---
title: "Final Project"
author: "Jordan Andrew, Pranshul Bhatnagar, Arushi Singh, and Jay Wu"
subtitle: "Due December 13th at 12:00pm"
format: pdf
editor: visual
---

#### Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(dplyr)
library(car)
library(ggplot2)
```

**Load the data and provide a `glimpse()`**:

```{r data-1, warning=FALSE}
library(dplyr)
music_data <- read.csv("data/mxmh_survey_results.csv")
glimpse(music_data)

```

# **Data Cleaning**

```{r}
# Showing which columns have null values and how many
music_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "na_count") %>%
  filter(na_count > 0) %>%
  arrange(desc(na_count))

```

```{r}
# Summary to spot outliers
summary(music_data$Age)

# Count unrealistic ages
sum(music_data$Age < 10 | music_data$Age > 90, na.rm = TRUE)

# View rows with unrealistic ages
music_data %>% 
  filter(Age < 10 | Age > 90)
```

```{r}

# Checking that music hours per day doesn't contain unrealistic # of hours
summary(music_data$Hours.per.day)

music_data %>% 
  filter(Hours.per.day < 0 | Hours.per.day > 16)

# we will remove the 89 year old who listens to rap exclusively while working and has no perceived mental health issues

```

```{r}
cols <- c("Anxiety", "Depression", "Insomnia", "OCD")

# Checking if any values for self reported mental illness are out of the 0-10 range
music_data %>% 
  summarise(across(all_of(cols), 
                   ~ sum(. < 0 | . > 10, na.rm = TRUE)))


```

```{r}
yn_cols <- c(
  "While.working",
  "Instrumentalist",
  "Composer",
  "Exploratory",
  "Foreign.languages"
)

# Checking to see if any of the "yes" "no" columns have values outside of these options
sapply(music_data[yn_cols], function(x) all(x %in% c("Yes", "No")))

lapply(music_data[yn_cols], function(x) setdiff(unique(x), c("Yes", "No")))

sapply(music_data[yn_cols], function(x) sum(x == "", na.rm = TRUE))

# We will drop the rows that having missing values here, as there is no clear way how to impute them


```

```{r}
freq_cols <- c(
  "Frequency..Classical.",
  "Frequency..Country.",
  "Frequency..EDM.",
  "Frequency..Folk.",
  "Frequency..Gospel.",
  "Frequency..Hip.hop.",
  "Frequency..Jazz.",
  "Frequency..K.pop.",
  "Frequency..Latin.",
  "Frequency..Lofi.",
  "Frequency..Metal.",
  "Frequency..Pop.",
  "Frequency..R.B.",
  "Frequency..Rap.",
  "Frequency..Rock.",
  "Frequency..Video.game.music."
)

allowed_freq <- c("Never", "Rarely", "Sometimes", "Very frequently")


# Checking to see if any of the frequency columns have values outside of "Never, Rarely, Sometimes, Very frequently"
sapply(music_data[freq_cols], function(x) sum(!(x %in% allowed_freq)))
```

```{r}
table(music_data$Music.effects)
```

# **General Data Cleaning**

```{r}
# Cleaning our data

# dropping BPM column
music_data <- music_data %>% select(-any_of("BPM"))

# dropping 1 row that has a null value for Age
music_data <- music_data %>% 
  filter(!is.na(Age))

# Dropping the 89 year old year old who listens to only rap while working 24 hours a day(unrealistic value)
music_data <- music_data %>%
  filter(!(Age == 89 & Hours.per.day == 24))

# Dropping the rows that have non "yes" "no" answers for questions that only permit those values (9 rows were removed here)
allowed_yn <- c("Yes", "No")

music_data <- music_data %>%
  filter(
    if_all(all_of(yn_cols), ~ .x %in% allowed_yn)
  )

# removing 7 rows where Music.effects has an empty string as a response
music_data <- music_data %>%
  filter(Music.effects != "")

# Keep only columns that exist in the data
freq_cols_present <- freq_cols[freq_cols %in% colnames(music_data)]

# Only rename if any of these columns exist
if (length(freq_cols_present) > 0) {
  music_data <- music_data %>%
    rename_with(.cols = all_of(freq_cols_present), 
                .fn = ~ gsub("Frequency\\.\\.", "freq_", .x) %>% 
                       gsub("\\.", "", .))  # Remove all remaining periods
}

music_data_clean <- music_data

glimpse(music_data_clean)

```

# **Research Question 1**

```{r}
# Question 1: How does the number of hours spent listening to music per day vary by depression status and primary streaming service?

# this will be the dataset used for our first research question
music_data_1 <- music_data_clean

unique(music_data_1$Primary.streaming.service)

music_data_1 %>%
  count(Primary.streaming.service)

# removing the row where Primary.streaming.service is empty
music_data_1 <- music_data_1 %>%
  filter(Primary.streaming.service != "")

music_data_1 %>%
  filter(Primary.streaming.service == "I do not use a streaming service.")

# removing the rows where participants stated they do not use a streaming service
music_data_1 <- music_data_1 %>%
  filter(Primary.streaming.service != "I do not use a streaming service.")

music_data_1 %>%
  count(Primary.streaming.service)
```

# **Modeling for Research Question 1**

```{r}
# Question 1: How does the number of hours spent listening to music per day vary by depression status and primary streaming service?

# fit our linear regression model
model_rq1_raw <- lm(
  Hours.per.day ~ Depression * Primary.streaming.service +
    Age +
    While.working +
    Instrumentalist +
    Composer +
    Exploratory +
    Foreign.languages +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects +
    freq_Classical +
    freq_Country +
    freq_EDM +
    freq_Folk +
    freq_Gospel +
    freq_Hiphop +
    freq_Jazz +
    freq_Kpop +
    freq_Latin +
    freq_Lofi +
    freq_Metal +
    freq_Pop +
    freq_RB +
    freq_Rap +
    freq_Rock +
    freq_Videogamemusic,
  data = music_data_1
)

summary(model_rq1_raw)

```

```{r}
plot(model_rq1_raw)
```
```{r}
# checking for multicollinearity
vif(model_rq1_raw, type = "predictor")
```




# **Model with Intentional/Relevant Predictors RQ1**

```{r}

# fit our linear regression model
model_rq1 <- lm(
  Hours.per.day ~ Depression * Primary.streaming.service +
    Age +
    While.working +
    Instrumentalist +
    Composer +
    Exploratory +
    Foreign.languages +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects,
  data = music_data_1
)

summary(model_rq1)

```

# **Diagnostics for Research Question 1 Model**

```{r}
plot(model_rq1)

```

```{r}
# checking for multicollinearity
vif(model_rq1, type = "predictor")
```

```{r}
p1 <- ggplot(music_data_1, aes(x = Hours.per.day)) +
geom_histogram(bins = 30, color = "black") +
labs(
x = "Hours of music listening per day",
y = "Count",
title = "Distribution of Hours.per.day"
)

p1
```

# **Log Transformation of Outcome Variable RQ1**

```{r}
# build log-transformed outcome
music_data_1 <- music_data_1 %>%
  mutate(log_hours = log(Hours.per.day + 0.1))

# run log-transform
model_rq1_log <- lm(
  log_hours ~ Depression * Primary.streaming.service +
    Age +
    While.working +
    Instrumentalist +
    Composer +
    Exploratory +
    Foreign.languages +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects,
  data = music_data_1
)

summary(model_rq1_log)
```


```{r}
# diagnostic plots for log-transformed model
plot(model_rq1_log)

# we can see that violations of homoscedasticity and normality have improved (no more strong violations)

```
```{r}
# Cook's distance for linear regression (RQ1)
cooks <- cooks.distance(model_rq1_log)

plot(cooks, type = "h",
     main = "Influential points measured by Cook’s distance",
     ylab = "Cook's distance",
     xlab = "Index")

# Identify influential observations
which(cooks > 0.04)

```
```{r}
# Removing the three main outliers shown in the diagnostic plots
music_data_1_subset <- music_data_1[!(rownames(music_data_1) %in% c("407", "345", "537")), ]

# Refitting the model
# run log-transform
model_rq1_log_final <- lm(
  log_hours ~ Depression * Primary.streaming.service +
    Age +
    While.working +
    Instrumentalist +
    Composer +
    Exploratory +
    Foreign.languages +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects,
  data = music_data_1_subset
)

summary(model_rq1_log_final)
```

```{r}
# Diagnostic plots for the final model, after log transformation and removal of outliers
plot(model_rq1_log_final)
```

# **Research Question 2**

```{r}
# Question 2: Does the number of hours spent listening to music each day increase the likelihood of reporting high depression symptoms?

# this will be the dataset used for our first research question
music_data_2 <- music_data_clean

# create high_depression as 1 if Depression >= threshold, 0 otherwise
music_data_2$high_depression <- ifelse(music_data_2$Depression >= 6, 1, 0)

# Fit logistic regression including all relevant predictors
model2 <- glm(
  high_depression ~ Hours.per.day +
    Age +
    Instrumentalist +
    Composer +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects,
  data = music_data_2,
  family = binomial
)

# View model summary
summary(model2)

```

```{r}
# Coefficients on the odds scale
exp(coef(model2))
```
# **Diagnostics RQ2**

# **Partial residual plot**

```{r}
# Partial residual plot for our main predictor (Hours.per.day)

# This checks linearity of Hours.per.day with log-odds.
crPlot(model2, "Hours.per.day")

```

# **Checking for multicollinearity**

```{r}
vif(model2)
```
# **Goodness-of-fit (Hosmer-Lemeshow test)** 

```{r}
# This test compares observed and expected frequencies of the outcome.
ResourceSelection::hoslem.test(model2$y, fitted(model2))

```

## Pearson residuals vs fitted plot##
```{r}
plot(model2, which = 1)
```

# **Checking for high leverage points**

```{r}
hatvalues <- hatvalues(model2)
which(hatvalues > (2*mean(hatvalues)))

```

# **Confusion Matrix** 

```{r}
probs <- predict(model2, type = "response")

pred_class <- ifelse(probs > 0.5, 1, 0)
pred_class <- factor(pred_class, levels = c(0,1))

caret::confusionMatrix(
  table(pred_class, music_data_2$high_depression),
  positive = "1",
  mode = "everything"
)

```

# **ROC Curve and AUC** 

```{r}
pROC::roc(music_data_2$high_depression, probs,
          print.thres = "best",
          print.auc = TRUE,
          legacy.axes = TRUE,
          plot = TRUE)

```

# **Updated Confusion Matrix with "best" Threshold**

```{r}
probs <- predict(model2, type = "response")

pred_class <- ifelse(probs > 0.456, 1, 0)
pred_class <- factor(pred_class, levels = c(0,1))

caret::confusionMatrix(
  table(pred_class, music_data_2$high_depression),
  positive = "1",
  mode = "everything"
)
```

# **Reduced Model Comparison (no Hours.per.day)**

```{r}

# reduced model without Hours.per.day
model2_red <- glm(
  high_depression ~
    Age +
    Instrumentalist +
    Composer +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects,
  data = music_data_2,
  family = binomial
)


# Deviance test
anova(model2_red, model2, test="Chisq")

AIC(model2_red)
AIC(model2)




```

# \*\* Confidence Intervals and Odds Ratios\*\*

```{r}
exp(cbind(OR = coef(model2), confint(model2)))

```
