---
title: "Final Project"
author: "Jordan Andrew, Pranshul Bhatnagar, Arushi Singh, and Jay Wu"
subtitle: "Due December 13th at 12:00pm"
format: pdf
editor: visual
---

#### Load Packages

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(dplyr)
library(car)
library(ggplot2)
```

**Load the data and provide a `glimpse()`**:

```{r data-1, warning=FALSE}
library(dplyr)
music_data <- read.csv("data/mxmh_survey_results.csv")
glimpse(music_data)

```

# **Data Cleaning**

```{r}
# Showing which columns have null values and how many
music_data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "na_count") %>%
  filter(na_count > 0) %>%
  arrange(desc(na_count))

```

```{r}
# Summary to spot outliers
summary(music_data$Age)

# Count unrealistic ages
sum(music_data$Age < 10 | music_data$Age > 90, na.rm = TRUE)

# View rows with unrealistic ages
music_data %>% 
  filter(Age < 10 | Age > 90)
```

```{r}

# Checking that music hours per day doesn't contain unrealistic # of hours
summary(music_data$Hours.per.day)

music_data %>% 
  filter(Hours.per.day < 0 | Hours.per.day > 16)

# we will remove the 89 year old who listens to rap exclusively while working and has no perceived mental health issues

```

```{r}
cols <- c("Anxiety", "Depression", "Insomnia", "OCD")

# Checking if any values for self reported mental illness are out of the 0-10 range
music_data %>% 
  summarise(across(all_of(cols), 
                   ~ sum(. < 0 | . > 10, na.rm = TRUE)))


```

```{r}
yn_cols <- c(
  "While.working",
  "Instrumentalist",
  "Composer",
  "Exploratory",
  "Foreign.languages"
)

# Checking to see if any of the "yes" "no" columns have values outside of these options
sapply(music_data[yn_cols], function(x) all(x %in% c("Yes", "No")))

lapply(music_data[yn_cols], function(x) setdiff(unique(x), c("Yes", "No")))

sapply(music_data[yn_cols], function(x) sum(x == "", na.rm = TRUE))

# We will drop the rows that having missing values here, as there is no clear way how to impute them


```

```{r}
freq_cols <- c(
  "Frequency..Classical.",
  "Frequency..Country.",
  "Frequency..EDM.",
  "Frequency..Folk.",
  "Frequency..Gospel.",
  "Frequency..Hip.hop.",
  "Frequency..Jazz.",
  "Frequency..K.pop.",
  "Frequency..Latin.",
  "Frequency..Lofi.",
  "Frequency..Metal.",
  "Frequency..Pop.",
  "Frequency..R.B.",
  "Frequency..Rap.",
  "Frequency..Rock.",
  "Frequency..Video.game.music."
)

allowed_freq <- c("Never", "Rarely", "Sometimes", "Very frequently")


# Checking to see if any of the frequency columns have values outside of "Never, Rarely, Sometimes, Very frequently"
sapply(music_data[freq_cols], function(x) sum(!(x %in% allowed_freq)))
```

```{r}
table(music_data$Music.effects)
```

# **General Data Cleaning**

```{r}
# Cleaning our data

# dropping BPM column
music_data <- music_data %>% select(-any_of("BPM"))

# dropping 1 row that has a null value for Age
music_data <- music_data %>% 
  filter(!is.na(Age))

# Dropping the 89 year old year old who listens to only rap while working 24 hours a day(unrealistic value)
music_data <- music_data %>%
  filter(!(Age == 89 & Hours.per.day == 24))

# Dropping the rows that have non "yes" "no" answers for questions that only permit those values (9 rows were removed here)
allowed_yn <- c("Yes", "No")

music_data <- music_data %>%
  filter(
    if_all(all_of(yn_cols), ~ .x %in% allowed_yn)
  )

# removing 7 rows where Music.effects has an empty string as a response
music_data <- music_data %>%
  filter(Music.effects != "")

# Keep only columns that exist in the data
freq_cols_present <- freq_cols[freq_cols %in% colnames(music_data)]

# Only rename if any of these columns exist
if (length(freq_cols_present) > 0) {
  music_data <- music_data %>%
    rename_with(.cols = all_of(freq_cols_present), 
                .fn = ~ gsub("Frequency\\.\\.", "freq_", .x) %>% 
                       gsub("\\.", "", .))  # Remove all remaining periods
}

music_data_clean <- music_data

glimpse(music_data_clean)

```

# **Research Question 1**

```{r}
# Question 1: How does the number of hours spent listening to music per day vary by depression status and primary streaming service?

# this will be the dataset used for our first research question
music_data_1 <- music_data_clean

unique(music_data_1$Primary.streaming.service)

music_data_1 %>%
  count(Primary.streaming.service)

# removing the row where Primary.streaming.service is empty
music_data_1 <- music_data_1 %>%
  filter(Primary.streaming.service != "")

music_data_1 %>%
  filter(Primary.streaming.service == "I do not use a streaming service.")

# removing the rows where participants stated they do not use a streaming service
music_data_1 <- music_data_1 %>%
  filter(Primary.streaming.service != "I do not use a streaming service.")

music_data_1 %>%
  count(Primary.streaming.service)
```

# **Modeling for Research Question 1**

```{r}
# Question 1: How does the number of hours spent listening to music per day vary by depression status and primary streaming service?

# fit our linear regression model
model_rq1 <- lm(
  Hours.per.day ~ Depression * Primary.streaming.service +
    Age +
    While.working +
    Instrumentalist +
    Composer +
    Exploratory +
    Foreign.languages +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects +
    freq_Classical +
    freq_Country +
    freq_EDM +
    freq_Folk +
    freq_Gospel +
    freq_Hiphop +
    freq_Jazz +
    freq_Kpop +
    freq_Latin +
    freq_Lofi +
    freq_Metal +
    freq_Pop +
    freq_RB +
    freq_Rap +
    freq_Rock +
    freq_Videogamemusic,
  data = music_data_1
)

summary(model_rq1)

```

# **Diagnostics for Research Question 1 Model**

```{r}
plot(model_rq1)

```

```{r}
# checking for multicollinearity
vif(model_rq1, type = "predictor")
```

```{r}
# Possible transformations here (maybe get rid of those outlier points)
```

# **Log Transformation of Outcome Variable RQ1**

```{r}
p1 <- ggplot(music_data_1, aes(x = Hours.per.day)) +
geom_histogram(bins = 30, color = "black") +
labs(
x = "Hours of music listening per day",
y = "Count",
title = "Distribution of Hours.per.day"
)

p2 <- ggplot(music_data_1, aes(x = log_hours)) +
geom_histogram(bins = 30, color = "black") +
labs(
x = "log(Hours.per.day + 0.1)",
y = "Count",
title = "Distribution of log-transformed Hours.per.day"
)

p1
p2
```

```{r}
# build log-transformed outcome
music_data_1 <- music_data_1 %>%
  mutate(log_hours = log(Hours.per.day + 0.1))

# run log-transform
model_rq1_log <- lm(
  log_hours ~ Depression * Primary.streaming.service +
    Age +
    While.working +
    Instrumentalist +
    Composer +
    Exploratory +
    Foreign.languages +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects +
    freq_Classical +
    freq_Country +
    freq_EDM +
    freq_Folk +
    freq_Gospel +
    freq_Hiphop +
    freq_Jazz +
    freq_Kpop +
    freq_Latin +
    freq_Lofi +
    freq_Metal +
    freq_Pop +
    freq_RB +
    freq_Rap +
    freq_Rock +
    freq_Videogamemusic,
  data = music_data_1
)

summary(model_rq1_log)

# plot again
plot(model_rq1_log)



```

```{r}
# checking for multicollinearity
vif(model_rq1_log, type = "predictor")
```

# **Results for RQ1**

```{r}
library(broom)
library(dplyr)
library(knitr)
library(kableExtra)

# --- tidy model output ---
tbl_rq1 <- tidy(model_rq1_log) %>% 
  select(term, estimate, std.error, statistic, p.value) %>% 
  mutate(
    estimate  = round(estimate, 2),
    std.error = round(std.error, 2),
    statistic = round(statistic, 2),
    p.value   = ifelse(p.value < 0.001, "<0.001",
                       sprintf("%.2f", round(p.value, 2)))
  ) %>%
  rename(
    Variable    = term,
    Estimate    = estimate,
    `Std Error` = std.error,   
    `t-value`   = statistic,   
    `p-value`   = p.value
  )

# --- print nice LaTeX-style table ---
kable(
  tbl_rq1,
  booktabs = TRUE,
  caption = "Linear Regression Model Summary for RQ1"
) %>%
  kable_styling(position = "center", full_width = FALSE)


```

```{r}
# Distribution of music listening time across primary streaming services
ggplot(music_data_1, aes(x = Primary.streaming.service,
y = Hours.per.day)) +
geom_boxplot() +
labs(
x = "Primary streaming service",
y = "Hours of music listening per day",
title = "Hours of music listening by primary streaming service"
) +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This boxplot compares the distribution of daily music listening hours across streaming platforms.\
Spotify and YouTube Music users show wider variability and more high-hour outliers, whereas Pandora users report consistently lower listening times.
This descriptive comparison provides initial evidence that listening behavior may differ by platform, motivating the inclusion of streaming service in the regression model.

```{r}
# Raw relationship between depression and music listening time across streaming platforms
ggplot(music_data_1,
aes(x = Depression,
y = Hours.per.day,
color = Primary.streaming.service)) +
geom_jitter(width = 0.2, height = 0.2, alpha = 0.5) +
geom_smooth(method = "lm", se = FALSE) +
labs(
x = "Depression score (0–10)",
y = "Hours of music listening per day",
color = "Streaming service",
title = "Depression vs. hours of music listening by streaming service"
)
```

This descriptive plot visualizes the raw data without adjusting for any covariates.\
It shows how music listening hours vary with depression scores across streaming services,\
providing an exploratory view of potential differences in slopes before fitting the statistical model.

```{r}
# Model-estimated interaction between depression and streaming platform on music listening time

# Create a sequence of Depression values across the observed range
dep_seq <- seq(
  from = min(music_data_1$Depression, na.rm = TRUE),
  to   = max(music_data_1$Depression, na.rm = TRUE),
  length.out = 50
)

# Identify representative values for covariates 
# (e.g., median for continuous variables, modal category for categorical variables)

age_med <- median(music_data_1$Age, na.rm = TRUE)

# Function to compute the mode for categorical variables
mode_char <- function(x) names(sort(table(x), decreasing = TRUE))[1]

# Modes for binary/categorical covariates
while_mode   <- mode_char(music_data_1$While.working)
instr_mode   <- mode_char(music_data_1$Instrumentalist)
comp_mode    <- mode_char(music_data_1$Composer)
expl_mode    <- mode_char(music_data_1$Exploratory)
lang_mode    <- mode_char(music_data_1$Foreign.languages)
genre_mode   <- mode_char(music_data_1$Fav.genre)
effect_mode  <- mode_char(music_data_1$Music.effects)

# Mode function for frequency variables
freq_mode <- function(var) mode_char(music_data_1[[var]])

# Extract factor levels for streaming service
stream_levels <- levels(factor(music_data_1$Primary.streaming.service))

# Construct prediction grid:
# Vary Depression and streaming service,
# while holding all other covariates at representative values
newdat <- expand.grid(
  Depression = dep_seq,
  Primary.streaming.service = stream_levels
) %>%
  mutate(
    Age = age_med,
    While.working   = while_mode,
    Instrumentalist = instr_mode,
    Composer        = comp_mode,
    Exploratory     = expl_mode,
    Foreign.languages = lang_mode,
    Anxiety   = mean(music_data_1$Anxiety, na.rm = TRUE),
    Insomnia  = mean(music_data_1$Insomnia, na.rm = TRUE),
    OCD       = mean(music_data_1$OCD, na.rm = TRUE),
    Fav.genre     = genre_mode,
    Music.effects = effect_mode,
    freq_Classical      = freq_mode("freq_Classical"),
    freq_Country        = freq_mode("freq_Country"),
    freq_EDM            = freq_mode("freq_EDM"),
    freq_Folk           = freq_mode("freq_Folk"),
    freq_Gospel         = freq_mode("freq_Gospel"),
    freq_Hiphop         = freq_mode("freq_Hiphop"),
    freq_Jazz           = freq_mode("freq_Jazz"),
    freq_Kpop           = freq_mode("freq_Kpop"),
    freq_Latin          = freq_mode("freq_Latin"),
    freq_Lofi           = freq_mode("freq_Lofi"),
    freq_Metal          = freq_mode("freq_Metal"),
    freq_Pop            = freq_mode("freq_Pop"),
    freq_RB             = freq_mode("freq_RB"),
    freq_Rap            = freq_mode("freq_Rap"),
    freq_Rock           = freq_mode("freq_Rock"),
    freq_Videogamemusic = freq_mode("freq_Videogamemusic")
  )

# Generate model-based predicted values (predicted log-hours)
newdat$pred_log_hours <- predict(model_rq1_log, newdata = newdat)

# Plot the interaction between Depression and streaming service
ggplot(newdat,
       aes(x = Depression,
           y = pred_log_hours,
           color = Primary.streaming.service)) +
  geom_line(size = 1) +
  labs(
    x = "Depression score (0–10)",
    y = "Predicted log-hours of music listening",
    color = "Streaming service",
    title = "Interaction between depression and streaming service on music listening time"
  )


```

This plot displays predicted values from the multivariable regression model, holding all other covariates constant.\
The lines represent the adjusted effect of depression for each streaming service, allowing direct interpretation of the statistical interaction tested in the model.

# **Research Question 2**

```{r}
# Question 2: “How does the number of hours spent listening to music each day impact the severity of reported depression symptoms?”

# Use cleaned dataset
music_data_2 <- music_data_clean

# Fit logistic regression including all relevant predictors
model2 <- glm(
  Depression ~ Hours.per.day +
    Age +
    While.working +
    Instrumentalist +
    Composer +
    Exploratory +
    Foreign.languages +
    Anxiety +
    Insomnia +
    OCD +
    Fav.genre +
    Music.effects +
    freq_Classical +
    freq_Country +
    freq_EDM +
    freq_Folk +
    freq_Gospel +
    freq_Hiphop +
    freq_Jazz +
    freq_Kpop +
    freq_Latin +
    freq_Lofi +
    freq_Metal +
    freq_Pop +
    freq_RB +
    freq_Rap +
    freq_Rock +
    freq_Videogamemusic,
  data = music_data_2,
  family = binomial
)

# View model summary
summary(model2)

# Coefficients on the odds scale
exp(coef(model2))




```

# **Diagnostics RQ2**
